---
title: "Bloomington salaries analysis"
output: 
---

This file was the result of a public records request. It is the salaries of city employees in Bloomington, Indiana.

```{r message=F}
# remember the first code chunk is used to bring in the packages you need for this script.
# today we'll be importing an excel file (Blooming Salaries.xlsx, found in Files on Canvas) so we need the readxl package

library(tidyverse)
library(readxl)
```

```{r}
# you can read the documentation on this function to see what the required and optional arguments are. Like read_csv(), it's one required argument is a path/file to import. 
?read_excel

# if you're working in your R project for this class, you only need to tell R to look into the data folder and look for the file "Bloomington Salaries.xlsx"):
salaries <- read_excel("data/Bloomington Salaries.xlsx")

```

### These are the questions I always ask of data as I'm assessing it, getting familiar with it, etc:

1. How many rows? 1200
2. What is one row? 
  We think it's one employee, but how can we tell? Look for duplicates names... 
  There's one duplicate for "Emily Herr", find out if this is two different people through reporting
```{r}
salaries %>% 
  count(first_name, last_name) %>% 
  filter(n > 1)

salaries %>% filter(last_name == "Herr" & first_name == "Emily")
```
3. Look at every column.
Notes: 
- total_comp has no NAs
- there are no rows with both hourly_rate and salary values

```{r}
# this will give you some helpful information about the numbers columns in particular. Note that total_comp has no nulls (NAs):
summary(salaries)

# you can also test whether total_comp has NAs using is.na(): 
salaries %>% filter(is.na(total_comp))
# (no rows are returned)

# we assume that each person has either an hourly rate or a salary. Test whether any person has values in both columns. To do this, use ! in front of is.na() to search for values that AREN'T NA:
salaries %>% filter(!is.na(hourly_rate) & !is.na(salary_2021))
# (no rows returned)

# The ! can be used to negate other things as well. For example, using != means "not equal to": 
salaries %>% filter(job_title != "Intern")
```

For categorical variables such as `department` and `job_title`, check unique values to see if there are misspellings, typos, etc:
```{r}
salaries %>% count(department)
salaries %>% count(job_title)

# both of these seem clean.
```

### Now to some analysis questions:

Who made the most in overtime? 
```{r}
salaries %>% arrange(desc(overtime_oncall))
# Looks like Jeffrey Rodgers made $41K in overtime, bringint his total comp up to $106K. Why? The data can't tell us why; maybe he's gaming the system, maybe the police are shortstaffed and he has to work nights and weekends. That's a reporting question. The data tells us the "what", not the "why".
```

How much did the city payout in overtime? 
```{r}
# because the overtime_oncall column has NA values in it, if we sum it we'll get NA returned. This is an IMPORTANT thing to know about R:
salaries %>% summarise(total_overtime = sum(overtime_oncall))

# if you want to apply a summary function (sum, mean, median, etc) to a numbers column that has NAs in it, you have to explicitly tell those functions to remove the NAs, using the na.rm argument:
salaries %>% summarise(total = sum(overtime_oncall, na.rm=TRUE))

# note that na.rm=TRUE goes inside the sum() function.
```







