---
title: "Cleaning practice"
output: 
---

LIBRARIES:

```{r}
library(tidyverse)
library(janitor)
library(lubridate)
library(readxl)
```

DATA (located in Files on Canvas):

```{r}
# transactions file from our earlier work with FEC Campaign Finance
transactions <- read_csv("data/transactions.csv")

# deaths file h/t Mary Jo Webster, data editor of the Minnesota Star Tribune
mn_deaths <- read_csv("data/MN_opiate_deaths.csv")

# a Census Bureau file of population by state for five years, 2018 - 2022
state_pop <- read_csv("data/state_pop.csv")
```

TASKS: `transactions`

1.  convert the transaction_dt to a true date column
2.  create a new column in the `transactions` table that is a 5-digit zip code (always the first 5 when there are 9 digits.)

TASKS: `mn_deaths`

1.  check if there are duplicates (using StateID & using names/dob)
```{r}
mn_deaths %>% count(FIRSTNAME, MIDDLENAME, LASTNAME, BIRTHDATE) %>% arrange(desc(n))

mn_deaths %>% count(STATEID) %>% arrange(desc(n))
```


2.  check / clean up HISPANICETHNICITY (into standardized values)
3.  convert BIRTDATE & DEATHDATE

TASKS: `state_pop`

1.  Reorganize the file so that you can calculate the percent change for each states population from 2018 to 2022. 
2.  Which state has increased the most? 
3.  Which states have decreased in population? 
4.  Where does Missouri rank? 
