---
title: "R Notebook"
output:
  html_document:
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---

This dataset is based off of Billboard's weekly Hot 100 songs:
<https://www.billboard.com/charts/hot-100/>

It is compiled by Christian McDonald out of UT-Austin:
<https://github.com/utdata/rwd-billboard-data>

You can download it from Canvas.

```{r message=F}
library(tidyverse)
```

```{r}
hot100 <- read_csv("data/hot-100-current.csv")
```

### Data Assessment

How many rows? 345,000 What is one row? One song for one week What's in
every column? chart_week: every Saturday last_week: some 0s and some NAs
0s mean the song was on the charts before but not last week NAs mean the
song had not been on the charts before chart_week: goes back to 1958,
and most recent date is this coming Saturday

```{r}
# remember that count() is very useful in assessing what's in columns:
hot100 %>% count(current_week) # values 1 - 100
hot100 %>% count(last_week) # values 1 - 100 plus 0 and NA

# look at rows where last_week is NA or 0, is it because it's the song's first week on the charts?
hot100 %>% filter(is.na(last_week) | last_week==0) %>% 
  count(wks_on_chart)
# no, because wks_on_chart is > 1 for a bunch of these...

# take a look at a few examples:
hot100 %>% 
  filter(is.na(last_week) | last_week==0) %>% 
  filter(wks_on_chart==2)

# look at the details of a particular song to see if we can figure out how it works:
hot100 %>% 
  filter(title=="Santa Claus Is Comin' To Town" & performer=="Jackson 5") %>% 
  arrange(chart_week)
# yes, 0 means it's been on the chart before, just not last week. NA is truly new to the charts.

# look at the date range (this is a base R function, so it can be used this way or within summarise()):
range(hot100$chart_week)
```

---- SEPT 18 -----

Before we go through the assignment questions, think for a moment about what this dataset represents: it's the top 100 songs of each week *according to Billboard.* Well is Billboard the arbiter of good music? How does Billboard decide which songs make rank? You can google that, and you might find this page: [https://www.billboard.com/pro/ask-billboard-how-does-the-hot-100-work/](https://www.billboard.com/pro/ask-billboard-how-does-the-hot-100-work/)

Which says: 

"'Generally speaking, our Hot 100 formula targets a ratio of sales (35-45%), airplay (30-40%) and streaming (20-30%).'

Of course, that’s an overall target for 100 songs each week. That mark can change week-to-week. This week, though, the Hot 100 breaks down in line with the formula’s intent: sales, 39%; airplay, 34%; and, streaming, 27%."

Helpful, but also a little vague. And variable. Still, any story we write about this data can just attribute a song's rank to Billboard. For example, this week's top song, "A Bar Song (Tipsy)" isn't just *the most popular song* in a story, it's *the most popular song according to Billboard.*

**THINK** about your data, where it came from, how it was collected. Now to the questions:

Which artist has appeared on the Hot100 list ("the chart") the most
times?

```{r}
#The two lines of code below do exactly the same thing:

hot100 %>% group_by(performer) %>% summarise(n = n()) %>% arrange(desc(n))

hot100 %>% count(performer) %>% arrange(desc(n))
```

Which song has spent the most time on the chart?

```{r}
# You can get the answer several ways, here are two. Also keep in mind that song titles are not unique! If you don't include `performer` in the count(), you will see the song "Stay" at the top of the list, but there are 16 performers who sing a song called "Stay."

hot100 %>% count(title,performer) %>% arrange(desc(n))

hot100 %>% arrange(desc(wks_on_chart))

```

Which artist has spent the most weeks at #1?

```{r}
hot100 %>% filter(current_week==1) %>% count(performer) %>% arrange(desc(n))

# followup: look at Mariah Carey's songs: How much of this fame is attributable to "All I Want for Christmas is You"?
hot100 %>% 
  filter(performer == "Mariah Carey" & current_week==1) %>% 
  count(title) %>% 
  arrange(desc(n))
```

Which artist has had the most individual songs?

```{r}
# introducing the n_distinct() function here, which counts *unique* or *distinct* values instead of just values.

hot100 %>% 
  group_by(performer) %>% 
  summarise(n = n_distinct(title)) %>% 
  arrange(desc(n))

# But are they actually unique songs? Let's take a look at all the distinct values in `title` for just Taylor Swift:
hot100 %>% 
  filter(performer=="Taylor Swift") %>% 
  group_by(title) %>% 
  summarise(min = min(chart_week), max = max(chart_week))

# There are duplicates, where one song was rerecorded and named "Taylor's Version. Are those unique songs though because they were rerecorded? That's a methodological decision we have to make. We have to define *unique*.
```

```{r}
# Another twist in the plot of this data is that sometimes performers feature / collaborite with / team up with other performers. 
# Using the function grepl() within filter() lets us look for a string of text inside a cell, rather than just doing an exact match:

hot100 %>% filter(grepl("Beyonce",performer,ignore.case=T)) %>% 
  count(performer) %>% arrange(desc(n))

# Taylor mostly does solo but there are some where she features another artist, or another artist features her... there's also a /, with, etc. A variety of options. What will we include when analyzing Taylor Swift? Another methodological decision.
```

What was the #1 song when you were born?

```{r}
hot100 %>% filter(chart_week=="1983-05-28")
```

Homework Assignment: 
Explore Hot100 data in search of a potential story. Let your curiosity guide you; ask questions; use code to try and answer them. Even if you can't get the answers, I want to see your efforts. I also want to see your journalistic thinking at play here: what might make for a good story? What are you interested in?

Please submit your .Rmd with your questions / notes / thoughts / conclusions and your code. 