---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

This dataset is based off of Billboard's weekly Hot 100 songs: https://www.billboard.com/charts/hot-100/

It is compiled by Christian McDonald out of UT-Austin: 
https://github.com/utdata/rwd-billboard-data

You can download it from Canvas.

```{r message=F}
library(tidyverse)
```

```{r}
hot100 <- read_csv("data/hot-100-current.csv")
```

### Data Assessment
How many rows? 345,000
What is one row? One song for one week
What's in every column? 
  chart_week: every Saturday
  last_week: some 0s and some NAs
    0s mean the song was on the charts before but not last week
    NAs mean the song had not been on the charts before
  chart_week: goes back to 1958, and most recent date is this coming Saturday 
```{r}
# remember that count() is very useful in assessing what's in columns:
hot100 %>% count(current_week) # values 1 - 100
hot100 %>% count(last_week) # values 1 - 100 plus 0 and NA

# look at rows where last_week is NA or 0, is it because it's the song's first week on the charts?
hot100 %>% filter(is.na(last_week) | last_week==0) %>% 
  count(wks_on_chart)
# no, because wks_on_chart is > 1 for a bunch of these...

# take a look at a few examples:
hot100 %>% 
  filter(is.na(last_week) | last_week==0) %>% 
  filter(wks_on_chart==2)

# look at the details of a particular song to see if we can figure out how it works:
hot100 %>% 
  filter(title=="Santa Claus Is Comin' To Town" & performer=="Jackson 5") %>% 
  arrange(chart_week)
# yes, 0 means it's been on the chart before, just not last week. NA is truly new to the charts.

# look at the date range (this is a base R function, so it can be used this way or within summarise()):
range(hot100$chart_week)
```













