---
title: "tidycensus in class"
output:
  html_document:
    df_print: paged
---

### Preparation

Make sure you have an [activated API key from the census](https://api.census.gov/data/key_signup.html). Install the `usethis` package (`install.packages("usethis")`) and then run `usethis::edit_r_environ()` in the console below. This will open up a system file called `.REnviron`. In it, type this line: CENSUS_API_KEY= Then paste your API key, that long string of letters and numbers in your email from the Census Bureau. You do not need to use quotes. For example: CENSUS_API_KEY=a2392kf12oifo12n3kj (That's not a real key). Save and close the .REnviron file Restart R by going to Session \> Restart R

```{r}
library(tidyverse)
library(tidycensus)
```

### Documentation

Here's the [documentation for tidycensus](https://walker-data.com/tidycensus/articles/basic-usage.html#geography-in-tidycensus-1)

Kyle Walker's documentation is good, it's worth consulting frequently. In particular check out:

-   [Basic usage](https://walker-data.com/tidycensus/articles/basic-usage.html)

-   [Margins of error](https://walker-data.com/tidycensus/articles/margins-of-error.html)

### Basic usage for decennial census

In the basic usage section Kyle has some example code near the top to pull median age by state from the 2020 decennial census, using the function `get_decennial()`:

```{r}
age20 <- get_decennial(geography = "state", 
                       variables = "P13_001N", 
                       year = 2020,
                       sumfile = "dhc")

head(age20)
```

### Basic usage for the American Community Survey

If you scroll down near the bottom of the page to the header that says "Working with the ACS", Kyle also has some example code for the get_acs() function:

```{r}
###LIZ: note that here Kyle changes the name of the variable in the table to "medincome" so we know what "B19013_001" actually is: median household income

vt <- get_acs(geography = "county", 
              variables = c(medincome = "B19013_001"), 
              state = "VT", 
              year = 2021)

vt
```

Great, but we don't care that much about Vermont. Change it to give us the same median household income for our state's counties:

```{r}
MO <- get_acs(geography = "county", 
              variables = c(medincome = "B19013_001"), 
              state = "MO", 
              year = 2022)

MO
```

### Finding variables for "detailed" tables (IDs that start with `B`)

To get data, you need to name the variables that you want from the Census data. In order to know what those variables are, it's helpful to reference a table of all variables for the Census's "Detailed Tables" (these are tables with IDs that usually start with B):

```{r}
v22 <- load_variables(2022, "acs5", cache = TRUE)

head(v22)
```

In this table `V22` the `name` column is the variable name and `label` describes the variable. `concept` is the table name and `geography` is the smallest available geography. See more about [geographies in the Basic Usage section](https://walker-data.com/tidycensus/articles/basic-usage.html#geography-in-tidycensus) of the documentation.

Let's say we want to know the population under 5: the number of people born since the beginning of the pandemic. First we need to identify the variable that contains this information. By skimming through the variables we say the "Under 5 years" variable, so do a search for all variables containing that phrase:

```{r}
v22 %>% filter(grepl("Under 5", label))
```

Note that there are a lot; many are broken down by sex. But there is at least one category that is just "Estimate!!Total:!!Under 5 years", so search for that instead.

```{r}
v22 %>% filter(grepl("Estimate!!Total:!!Under 5 years", label))

# The variable we want is B06001_002. To look at Columbia, we'll use the "place" geography:
get_acs(geography = "place", 
              variables = c(under5 = "B06001_002"), 
              state = "MO", 
              year = 2022)
```

That got us the estimates; keep an eye on the margins of error (moe). If the moe is less than 1% of the estimate (`moe/estimate <= .01`), no problem. If it's between 1% and 10%, be cautious. If it's greater than 10%, the estimate is probably unreliable.

In order to understand these numbers, we really need denominators to compare different places in Missouri that have very different populations. Every Census table has a base population number that you should use as a denominator. I believe it's always the first variable in the table (`_001`). You can see all the variables in any given table by using `filter()` and `grepl()` again:

```{r}
v22 %>% filter(grepl("B06001_", name))
```

Now add total population to our request:

```{r}
under5_original <- get_acs(geography = "place", 
              variables = c(under5 = "B06001_002", total = "B06001_001"), 
              state = "MO", 
              year = 2022)
```

### Reshaping the results

It's very hard to make calculations between rows (such as dividing the under5 population by the total population for each place), so we'll reshape our data using pivot_wider:

```{r}
under5_original %>% 
  pivot_wider(names_from = "variable", values_from = c("estimate","moe"))
```

Then filter out any places where the under5 moe is more than 10% of the estimate:

```{r}
under5_original %>% 
  pivot_wider(names_from = "variable", values_from = c("estimate","moe")) %>% 
  filter(moe_under5/estimate_under5 <= .10)
```

Only 8 cities are left. Then calculate the under5 pop as a pct of the total and sort by that percentage:

```{r}
under5_original %>% 
  pivot_wider(names_from = "variable", values_from = c("estimate","moe")) %>% 
  filter(moe_under5/estimate_under5 <= .10) %>% 
  mutate(under5_pct = estimate_under5/estimate_total) %>% 
  arrange(desc(under5_pct))
```

What further comparisons could we make? Try comparing to the US as a whole; the basic request is the same, but change the geography (and remove the state argument):

```{r}
get_acs(geography = "us", 
              variables = c(under5 = "B06001_002", total = "B06001_001"), 
              year = 2022)

# easiest way to get the US percentage is just to divide the numbers using a calculator or R:
19004925/331097593
```

Next request: compare all places, not just the ones in Missouri. Copy the code above that pulls these variables by place and remove the state argument:

```{r}
get_acs(geography = "place", 
              variables = c(under5 = "B06001_002", total = "B06001_001"), 
              year = 2022) %>% 
  pivot_wider(names_from = "variable", values_from = c("estimate","moe")) %>% 
  filter(moe_under5/estimate_under5 <= .10) %>% 
  mutate(under5_pct = estimate_under5/estimate_total) %>% 
  arrange(desc(under5_pct))
```

Another way to find variables is to use [data.census.gov](https://data.census.gov); search for a topic you're interested in, such as "internet". The websit will give you a list of tables that match your search; explore them a little and see which one has the information you need. Then grab that table's ID. In this case, the table "S2801" has what we want (households without an internet subscription, namely). 

### Finding variables for "subject" tables (IDs that start with `S`)

```{r}
# add the suffix "/subject" to the second argument of the load_variables() function:
S22 <- load_variables(2022, "acs5/subject", cache = TRUE)
```

Now that you've got all the subject table variables, use `filter()` and `grepl()` again to find all the variables in that table:
```{r}
internet <- S22 %>% filter(grepl("S2801_", name))
```

Search that table for the estimate of households without an internet connection: "S2801_C01_019". 
Remember that the first variable will usually be our denominator. In this case we aren't measuring population (or people), we're measuring households. There are three main units of measurement in the Census: people, households and families. You can look up the specific definitions (and differences) between households and families on your own.

Now, what pct of Columbia households don't have an internet subscription?
```{r}
no_internet <- get_acs(geography = "place", 
              variables = c(none = "S2801_C01_019", total = "S2801_C01_001"), 
              state = "MO", 
              year = 2022)

no_internet %>% 
  pivot_wider(names_from = "variable", values_from = c("estimate","moe")) %>% 
  filter(estimate_total >= 5000) %>% 
  mutate(none_pct = estimate_none/estimate_total) %>% 
  arrange(desc(none_pct ))

# beware that the moe for Columbia is more than 10% of the estimate!
```
