---
title: "Weather events analysis"
output:
---

The data for this analysis can be found on the NOAA's Storm Prediction Center webpage: 
[https://www.spc.noaa.gov/wcm/#data](https://www.spc.noaa.gov/wcm/#data)


Start with the packages you need for this script:
```{r message=F}
library(tidyverse)
```

Load some data
```{r}
# this will spit the table out into the console below the code chunk:
read_csv("data/weather_events.csv")

# this will save the table in a variable called "events" that we can work with later:
events <- read_csv("data/weather_events.csv")
```

Assess the data: one of the most important steps in data analysis is getting to know your data, figuring out where it is messy, what's missing, looking for areas where it needs to be cleaned, etc. 

Good places to start: 
-   How many rows?
-   What is one row? 
-   What's in every column?
-   What are the boundaries?
```{r}
# what's the earliest and latest year? 
# the range() function gives the min and max values for any column:
range(events$year)

# you can use the summary function to see basic summary statistics on every column (most useful for numbers columns:)
summary(events)
```

Preview of tidyverse functions:
```{r}
# the select() function returns particular columns:
events %>% select(date, time, type)

# the arrange() function sorts your data by some column (or columns):
events %>% arrange(date)

# to see the sort in reverse (descending), add the desc() function inside the arrange() function:
events %>% arrange(desc(date))

# you can pipe your data through multiple functions:
events %>% select(date,time,type,state) %>% arrange(desc(date))

# the filter() function shows you a subset of rows based on some criteria:
events %>% filter(state == "MO")

# show these same rows but sorted by date
events %>% filter(state == "MO") %>% arrange(desc(date))

# the count() function gives you a distribution summary for any column:
events %>% count(type)
```

To read the documentation for a function, type the function name preceded by a question mark:
```{r}
?read_csv
```

More about filter(): adding multiple criteria, separated by either & (and) or | (or)
```{r}
# looking for only tornadoes in Missouri:
events %>% filter(state=="MO" & type=="tornado")

# looking for only tornadoes in Missouri that resulted in injuries (injuries is a numbers column, numbers do not go in double quotes):
events %>% filter(state=="MO" & type=="tornado" & injuries > 0)

# string these together to find out which tornado in Missouri resulted in the most injuries
events %>% 
  filter(state=="MO" & 
           type=="tornado" & 
           injuries > 0) %>% 
  arrange(desc(injuries))

# how does this compare to tornadoes in the entire country?
events %>% 
  filter(type=="tornado" & 
           injuries > 0) %>% 
  arrange(desc(injuries)) 
  
# use | to broaden your results - find all events in Missouri or Kansas:
events %>% filter(state=="MO" | state=="KS")
```

Overview of summarise(): use it to create summary statistics from your dataset. Choose which summary function (sum, median, mean, etc) you want to use:
```{r}
# this returns a sum of all fatalities from the dataset:
events %>% 
  summarise(sum(fatalities))

# add a filter to see how many of those fatalities were in Missouri: 
events %>% 
  filter(state=="MO") %>% 
  summarise(sum(fatalities))

# best practice is to name the resulting columns that come out of summarise():
events %>% 
  summarise(total = sum(fatalities))
```

Pair summarise() with group_by() to analyze groups within your data
```{r}
# to compare a sum of fatalities by state, first use group_by() so that summarise() acts on groups rather than the
events %>% 
  group_by(state) %>% 
  summarise(total = sum(fatalities))

# add arrange() to see which state had the most fatalities overall
events %>% 
  group_by(state) %>% 
  summarise(total = sum(fatalities)) %>% 
  arrange(desc(total))

# remember the count() function does something similar: provides a count of rows per unique value in any column:
events %>% count(state)

# to reproduce this using group_by() and summarise(), use the n() summary function which counts rows. This produces the exact same result as the code above:
events %>% 
  group_by(state) %>% 
  summarise(n = n())

# the benefit of using group_by() and summarise() is adding additional summary functions to your analysis:
events %>% 
  filter(type=="tornado") %>% 
  group_by(state) %>% 
  summarise(n = n(), total_fatal = sum(fatalities)) 

# add in additional functions to find out which state had the most fatalities by tornado, and how many states had at least one hundred fatalities (remember, order matters here):
events %>% 
  filter(type=="tornado") %>% 
  group_by(state) %>% 
  summarise(n = n(), 
            total_fatal = sum(fatalities), 
            total_inj = sum(injuries),
            avg_fata = mean(fatalities)) %>% 
  arrange(desc(total_fatal)) %>% 
  filter(total_fatal > 100)

# if you want to continue analyzing just the tornadoes, create a new table:
tornadoes <- events %>% filter(type=="tornado")
```


















